Capturing regeneration in models of forest dynamics

### 2022-11-16 manuscript/manuscriptRW.Rmd
manuscript: added most of the coments from Harald on the results bullet points&further changes in the results section

Regeneration levels, species composition and mortality between
recruitment diameters thresholds of 7 and 10cm varied across model
simulation results and in comparison with the observations. As expected,
recruitment was overestimated across most models for both the 7 and 10
cm diameter thresholds, with the exception of SYBILA, Landis II and
aDGVM2 that estimated recruitment rates at the lower end of the
plausibility interval of the empirical data (Figure \@ref(fig:over)).
The models with highest overestimation were PICUS and TreeMig. For most
of the models, the variability across sites in recruitment levels
(visible from the IQR in the box plots of Figure \@ref(fig:over)) is
similar to or smaller than the observed variability for most of the
models, with the exception of PICUS, where the mean recruitment
variability is much larger.

@@

Models did not match the observed patterns in recruitment across soil
moisture and temperature gradients and there was either a clear common
pattern across models. Looking at the soil moisture gradients (FigureS
2), we can see that a group of models decreases recruitment with
increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS
and LandClim), while few models have the opposite trend (ForCEEPs(f),
FORMIND and LPJ-GUESS), and the rest of do not have a clear pattern
(ForCEEPS, iLand, Landis II, TreeMig and aDGVM2). Across temperature
gradients all the models that had a decrease in recruitment with
increasing soil moisture showed the opposite trend (more recruitment
with increasing temperature) except for 4C (FigureS 3). The rest of the
models did not have a clear patterns across the temperature gradient.

@@

Most models match the level of diversity (as measured by the Shannon
index) of the observed data quite well (Figure \@ref(fig:H710)),
although they consistently underestimate its variability across sites.
Four models overestimate recruitment diversity (4C, PICUS, LANDIS-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case as this model simulated 5 species only. In most
models there are small differences across sites regarding the species
richness of recruitment between 7 and 10 cm. Most models maintain or
even increase in recruitment diversity between the 7 and 10 cm
thresholds, and the same is visible from the empirical data, however the
differences are not significant EXPLAIN WHY  "Can we understand this (I
know this is for the Discussion, but still...)?? This cannot be species
richness (as there are no new species popping between 7 and 10 cm DBH).
I'd also expect the number of species to be lower at 10 cm than at 7 cm
(because some will die off due to insufficient light etc.), but this is
not reflected in H' because species with zero abundance are ignored
(ln[0] is not defined). So a strongly increasing evenness must be
driving this increase of H'."

@@

### **Looking at the recruitment share of the 5 main tree species we found that**

-   Most of the models capture right the share of basal area recruitment
    of the 5 main tree species compared to the observed data in the
    environmental gradient (Figure \@ref(fig:diff)).

-   The models differ in the way they represent the climate niche of
    individual main species (Figure \@ref(fig:diff)).

-   Some models overestimated across the whole gradient one of these 5
    main tree species (for example 4C ForCEEPS, ForCEEPS(f), LandClim,
    FORMIND) while other models overestimate more than one species such
    as Landis II, xComp, PICUS or SYBILA (Figure \@ref(fig:diff)).

-   *Abies alba* was well captured across the gradient for most of the
    models except for xComp, Picus, LanClim and LandisII that
    over-estimated the recruitment share difference. There are also some
    particularities from selected models such as the case of 4C that did
    not simulate *Abies alba* (Figure \@ref(fig:diff)).

-   *Fagus sylvatica* was mainly underestimated cross the environmental
    gradient, with the exception of 4C that overestimate this species
    across the whole environmental gradient (Figure \@ref(fig:diff)).

-   Most of the models over-estimated *Picea abies* recruitment share on
    colder sites (Figure \@ref(fig:diff)).

-   *Pinus sylvestris* and *Quercus* spp. are well capture for those
    models that have recruitment in these species, but the recruitment
    values are always lower so the range of variation is also always
    around 0 (to be confirmed!). However some models lack of recruitment
    for these species (Figure \@ref(fig:diff)).

-   The values that we are seeing are means of sample per site and mean
    across the sites included in each environmental bin. It is important
    to noticed that those groups of sites in the empirical data where
    the recruited basal area of the species was 0 and the stand
    recruited basal area for all species was also 0, the cells color is
    represented as grey (Figure \@ref(fig:diff)). The lack of value in
    the empirical data can be originated due to not recruitment in the
    group of sites or that the sites did not have recruitment values for
    7cm threshold.

    Shall we plot (Figure \@ref(fig:diff)) for 10cm in the appendix and
    remove FiguresS 10-12?

```{r diff, echo=FALSE, fig.cap="Departure of each model in the representation of the main species recruitment relative abundance from the observations across the environmental gradients for a 7cm treshold. The values shown are the mean of samples per site and across sites in bins, where the range of the climatic water balance (cm) and seasonal degree-day sum (dºC) was divided in 10 bins. The point sizes represent ratio between recruited basal area of the species and the total recruited basal area for all the species. The color gradient shows the difference between the simulated and observed ratio between recruited basal area of the species and the total recruited basal area: grey cell represent areas of the envirnomental gradient in which the model had R BA share  and the empirical dara did not becuase both recruitment and stand level recruitments were equalt to zero.", out.width='100%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles.png"))
```

>>>>

## Recruitment levels

Results related to regeneration levels, species composition and
mortality in recruitment varied across models and in comparison with the
observations. As expected, recruitment was overestimated across most
models for both the 7 and 10 cm diameter thresholds, with the exception
of SYBILA, Landis II and aDGVM2 that estimated recruitment rates at the
lower end of the plausibility interval of the empirical data (Figure
\@ref(fig:over)). The models with highest overestimation were PICUS and
TreeMig. For most of the models, the variability across sites in
recruitment levels (visible from the interquartile range in the box
plots of Figure \@ref(fig:over)) is similar to or smaller than the
observed variability for most of the models, with the exception of
PICUS, where the mean recruitment variability is much larger.

@@

Models did not match the observed patterns in recruitment across soil
moisture and temperature gradients and there was either a clear common
pattern across models. Looking at the soil moisture gradients (FigureS
2), we can see that a group of models decreases recruitment with
increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS
and LandClim), while few models had the opposite trend (ForCEEPs(f),
FORMIND and LPJ-GUESS). Across temperature gradients all the models that
had a decrease in recruitment with increasing soil moisture showed the
opposite trend (more recruitment with increasing temperature) except for
4C (FigureS 3). The rest of the models did not have a clear pattern
across the temperature gradient.

@@

## Species composition in recruitment

Most of the models match the level of diversity (as measured by the
Shannon index) of the observed data quite well (Figure \@ref(fig:H710)),
although they consistently underestimate its variability across sites.
Four models overestimate recruitment diversity (4C, PICUS, LANDIS-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case as this model simulated 5 species only. In most
models there are small differences across sites regarding the species
richness of recruitment between 7 and 10 cm. Most models maintain or
even increase in recruitment diversity between the 7 and 10 cm
thresholds, and the same is visible from the empirical data, however the
differences are not significant. **EXPLAIN WHY  "Can we understand this
(I know this is for the Discussion, but still...)?? This cannot be
species richness (as there are no new species popping between 7 and 10
cm DBH). I'd also expect the number of species to be lower at 10 cm than
at 7 cm (because some will die off due to insufficient light etc.), but
this is not reflected in H' because species with zero abundance are
ignored (ln[0] is not defined). So a strongly increasing evenness must
be driving this increase of H'."**

@@

Silver fir (*Abies alba*), beech (*Fagus sylvatica*), Scots pine (*Pinus
sylvestris*) and oaks (*Quercus* spp) were the most frequent species in
the observed data. Most models capture the share of basal area of
recruitment of the five main tree species compared to the observed data
along the two major gradients of climate (Figure \@ref(fig:diff) and
FigureS XX). However, the models differ in the way they represent the
climate niche of individual main species. First, some models
overestimated recruitment across the whole gradient for one of the 5
main tree species (for example, ForCEEPS, ForCEEPS(f),LandClim or
LPJ-GUESS), while other models overestimated recruitment for more than
one species for larger parts of the climate gradient, such as xComp,
PICUS , SYBILA or LPJ-GUESS (cf. red colors in Figure \@ref(fig:diff)).
This trends were consistent with those when the regeneration thresholds
was at 7 and 10cm although the exact changes in recruited basal area
share changed (FigureS XX).

When looking individually to each of the five main tree species it is
possible to observe differences in the simulated recruitment results.
Silver fir (*Abies alba*) was well captured across the gradient by most
models except for xComp, Picus, LandClim and Landis-II, which
overestimated its recruitment share. There are also some particularities
evident for some models, such as 4C that did not simulate *Abies alba*.
*Fagus sylvatica* was underestimated by many models across the
environmental gradient, including ForCEEPS(f) and FORMIND that did not
have regeneration in most of the sites for this species. Most of the
models over-estimated *Picea abies* recruitment share on colder sites.
*Pinus sylvestris* and *Quercus* spp. represented a smaller share of the
recruited species in the observed data and localized on certain areas of
the environmental gradient, however many models did have recruitment for
these two species in most of the climatic gradient although at basal
area share. Five models (4C, FORMIND, SYBILA, xComp, LandClim) had
almost no generation or did not regenerate *Pinus sylvestris* and
*Quercus* spp.

It is important to keep in mind that the values shown in Figure
\@ref(fig:diff) are the means of all samples per site, and the mean
across the sites in each environmental bin. In this figure we are seeing
together two types of zeros (represented with a lack of dots in the
corresponding tile), the first type of zero are those sites where the
recruited basal area of the species was zero and the stand recruited
basal area for all species was also zero, those were considered as zero
BA share; the second type of zeros are those sites where there was no
recruitment for that species but there was recruitment for other
species.

```{r diff, echo=FALSE, fig.cap="Departure of each model in the representation of the main species recruitment relative abundance from the observations across the environmental gradients for a 7cm treshold. The values shown are the mean of samples per site and across sites in bins, where the range of the climatic water balance (cm) and seasonal degree-day sum (dºC) was divided in 10 bins. The point sizes represent ratio between recruited basal area of the species and the total recruited basal area for all the species. The color gradient shows the difference between the simulated and observed ratio between recruited basal area of the species and the total recruited basal area: grey cell represent areas of the envirnomental gradient in which the model had R BA share  and the empirical dara did not becuase both recruitment and stand level recruitments were equal to zero.", out.width='100%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles_7.png"))
```
